---
title: "ACLED Codebook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACLED Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Pages under the ACLED Codebook tab describe the variables in ACLED's data and provide usage examples. Users should consult the complete guide summarizing ACLED's methodology at [here](https://acleddata.com/download/2827/) for more details on each column.

# ACLED Overview

The ACLED project codes reported information on the type, agents, location, date, and other characteristics of political violence events, demonstrations and politically relevant non-violent events (referred to as 'Strategic developments'). ACLED focuses on tracking a range of violent and non-violent actions by political agents, including governments, rebels, militias, identity groups, political parties, external actors, rioters, protesters and civilians.

Political violence is defined as the use of force by a group with a political purpose or motivation. ACLED records political violence through its constituent events, the intent of which is to produce a comprehensive overview of all forms of political disorder, expressed through violence and demonstrations, within and across states. A politically violent event is a single altercation where often force is used by one or more groups toward a political end, although some non-violent instances -- including protests and strategic developments -- are included in the dataset to capture the potential pre-cursors or critical junctures of a violent conflict.

The fundamental unit of observation in ACLED is the *event*. Events involve designated actors -- e.g. a named rebel group, a militia or state forces. They occur at a specific named location (identified by name and geographic coordinates) and on a specific day.

## Column List

```{r, echo = FALSE, warning=F, message=FALSE}
library(kableExtra)
library(dplyr)


returned_data_acled <- data.frame(`Column Name` = c("event_id_cnty", "event_date", "year", "time_precision", "disorder_type",
                                    "event_type", "sub_event_type", "actor1", "assoc_actor_1", "inter1",
                                    "actor2", "assoc_actor_2", "inter2", "interaction", "civilian_targeting",
                                    "iso", "region", "country", "admin1", "admin2", "admin3", "location",
                                    "latitude", "longitude", "geo_precision", "source", "source_scale", "notes",
                                    "fatalities", "tags", "timestamp"),
                 Type = c("string", "date", "double", "double", "string", "string", "string", "string", "string", "double",
                          "string", "string", "double", "double", "string", "double", "string", "string", "string", "string",
                          "string", "string", "double", "double", "double", "string", "string", "string", "double",
                          "string", "double"),
                 Description = c("A unique alphanumeric event identifier by number and country acronym. This identifier remains constant even when the event details are updated.",
                                 "The date on which the event took
place. Recorded as Year-Month-Day.", "The year in which the event took
place.",
                                 "A numeric code between 1 and 3
indicating the level of precision of
the date recorded for the event.
The higher the number, the lower
the precision.",
                                 "The disorder category an event
belongs to.
",
                                 "The type of event; further
specifies the nature of the event.", "A subcategory of the event type.", "One of two main actors involved
in the event (does not necessarily
indicate the aggressor).",
                                 "Actor(s) involved in the event
alongside ‘Actor 1’ or actor
designations that further identify
‘Actor 1’", "A numeric code between 0 and
8 indicating the type of ‘Actor 1’.
",
                                 "One of two main actors involved
in the event (does not necessarily
indicate the target or victim).", "Actor(s) involved in the event
alongside ‘Actor 2’ or actor
designation further identifying
‘Actor 2’. ",
                                 "A numeric code between 0 to 8
indicating the type of ‘Actor 2’.", "A two-digit numeric code
(combination of ‘Inter 1’ and ‘Inter
2’) indicating the two actor types
interacting in the event.",
                                 "This column indicates whether
the event involved civilian
targeting. ", "A unique three-digit numeric
code assigned to each country or
territory according to ISO 3166.",
                                 "The region of the world where the
event took place.", "The country or territory in which
the event took place.",
                                 "The largest sub-national
administrative region in which the
event took place.",
                                 "The second largest sub-national administrative region in which the event took place.",
                                 "The third largest sub-national administrative region in which the event took place.",
                                 "The name of the location at which
the event took place.", "The latitude of the location in
four decimal degrees notation
(EPSG:3857).", "The longitude of the location in
four decimal degrees notation
(EPSG:3857).",
                                 "A numeric code between 1 and
3 indicating the level of certainty
of the location recorded for the
event. The higher the number,
the lower the precision.",
                                 "The sources used to record the
event. Separated by a semicolon.", "An indication of the geographic
closeness of the used sources to
the event", "A short description of the event.",
                                 "The number of reported fatalities
arising from an event. When there
are conflicting reports, the most
conservative estimate is recorded", "Additional structured information
about the event. Separated by a
semicolon.",
                                 "An automatically generated Unix timestamp that represents the exact date and time an event was last uploaded to the ACLED API."), check.names = F
)

returned_data_acled %>% 
  kable(align=rep('c')) %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(2, width = "20em") %>% 
  column_spec(3, width = "50em")

```

## ACLED Help

To better understand what each column in our dataset represents, `acledR::acled_help()` redirects you to relevant documentation for the column in question. 

```{r,eval=FALSE}

acled_help(column = NULL)

```

The function is very straightforward, only requiring one argument.  `column` refers to the column which users may be interested on exploring further. 

**Note**: `acled_help()` only supports one column at a time. 

For example, if you would like to know more about the *fatalities* column, you can write `acled_help(column="fatalities")` which will prompt a fatalities vignette in the RStudio Help pane.

While some columns have an standalone documentation vignette, e.g. Fatalities; others may be part of another vignette, e.g. admins are in the geographical vignette. 
