---
title: "ACLED Codebook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACLED Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=F, message=F,
  echo = F,
  cache = T
)
```

Pages under the ACLED Codebook tab describe the variables in the ACLED data and shows provides usage examples via the `acledR` package. Users should consult the complete guide summarizing ACLED's methodology at [here](https://acleddata.com/acleddatanew//wp-content/uploads/2021/11/ACLED_Codebook_v1_January-2021.pdf) for more details on each variable.

# ACLED Overview

The ACLED project codes reported information on the type, agents, location, date, and other characteristics of political violence events, demonstrations and select politically relevant non-violent events. ACLED focuses on tracking a range of violent and non-violent actions by political agents, including governments, rebels, militias, identity groups, political parties, external actors, rioters, protesters and civilians.

Political violence is defined as the use of force by a group with a political purpose or motivation. ACLED records political violence through its constituent events, the intent of which is to produce a comprehensive overview of all forms of political disorder, expressed through violence and demonstrations, within and across states. A politically violent event is a single altercation where often force is used by one or more groups toward a political end, although some non-violent instances -- including protests and strategic developments -- are included in the dataset to capture the potential pre-cursors or critical junctures of a violent conflict.

The fundamental unit of observation in ACLED is the *event*. Events involve designated actors -- e.g. a named rebel group, a militia or state forces. They occur at a specific named location (identified by name and geographic coordinates) and on a specific day.

## Variable List {#variable-list}

```{r}
library(kableExtra)
library(dplyr)
acledR::acled_codebook %>% 
  kable() %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(2, width = "20em") %>% 
  column_spec(3, width = "50em")
```

## ACLED Help

To further assist our users to understand what each column in our dataset represents, we have added a function that redirects users to relevant documentation for the column in question. That function is `acled_help()`:

```{r,eval = FALSE}

acled_help(column = NULL)

```

As users can observe, the function is relatively straightforward, only requiring one argument. In this case `column` refers to the column which users may be interested on exploring further. Note: `acled_help()` only supports one column at a time. 

For example, if an user would like to know more about the fatalities column, he/she/they can write `acled_help(column="fatalities")` which will prompt the fatalities vignette.

Some columns have an standalone documentation vignette, e.g. Fatalities. However, others will be either part of another vignette, e.g. admins and geographical vignette.


