---
title: "ACLED Codebook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACLED Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Pages in the ACLED Codebook tab describe the variables found in ACLED's data, while also providing examples of their use. You can consult the complete guide summarizing ACLED's methodology [here](https://acleddata.com/download/2827/) for a more thorough description of the methodology underlying each column.

# ACLED Overview

The ACLED project codes and standardizes reported information on the type, agents, location, date, and other characteristics of events of political violence, demonstrations, and politically relevant non-violent events (referred to as 'Strategic developments'). ACLED focuses on tracking a range of violent and non-violent actions by political agents, including governments, rebels, militias, identity groups, political parties, external actors, rioters, protesters and civilians.


The fundamental unit of observation in ACLED is the *event*. Events involve designated actors -- e.g. a named rebel group, a militia or state forces. They occur at a specific named location (identified by name and geographic coordinates) and on a specific day. For more information on what qualifies as an event, and how ACLED codes long-running conflicts, please see our [knowledge base](https://acleddata.com/knowledge-base/how-are-events-disaggregated/). Every event is described by information in a standardized set of columns comprising ACLED’s core dataset. Below you can find a brief description of each column. 

## Column List

```{r, echo = FALSE, warning=F, message=FALSE}
library(kableExtra)
library(dplyr)


returned_data_acled <- data.frame(`Column Name` = c("event_id_cnty", "event_date", "year", "time_precision", "disorder_type",
                                    "event_type", "sub_event_type", "actor1", "assoc_actor_1", "inter1",
                                    "actor2", "assoc_actor_2", "inter2", "interaction", "civilian_targeting",
                                    "iso", "region", "country", "admin1", "admin2", "admin3", "location",
                                    "latitude", "longitude", "geo_precision", "source", "source_scale", "notes",
                                    "fatalities", "tags", "timestamp"),
                 Type = c("string", "date", "double", "double", "string", "string", "string", "string", "string", "double",
                          "string", "string", "double", "double", "string", "double", "string", "string", "string", "string",
                          "string", "string", "double", "double", "double", "string", "string", "string", "double",
                          "string", "double"),
                 Description = c("A unique alphanumeric event identifier by number and country acronym. This identifier remains constant even when the event details are updated.",
                                 "The date on which the event took
place. Recorded as Year-Month-Day.", "The year in which the event took
place.",
                                 "A numeric code between 1 and 3
indicating the level of precision of
the date recorded for the event.
The higher the number, the lower
the precision.",
                                 "The disorder category an event
belongs to.
",
                                 "The type of event; further
specifies the nature of the event.", "A subcategory of the event type.", "One of two main actors involved
in the event (does not necessarily
indicate the aggressor).",
                                 "Actor(s) involved in the event
alongside ‘Actor 1’ or actor
designations that further identify
‘Actor 1’", "A numeric code between 0 and
8 indicating the type of ‘Actor 1’.
",
                                 "One of two main actors involved
in the event (does not necessarily
indicate the target or victim).", "Actor(s) involved in the event
alongside ‘Actor 2’ or actor
designation further identifying
‘Actor 2’. ",
                                 "A numeric code between 0 to 8
indicating the type of ‘Actor 2’.", "A two-digit numeric code
(combination of ‘Inter 1’ and ‘Inter
2’) indicating the two actor types
interacting in the event.",
                                 "This column indicates whether
the event involved civilian
targeting. ", "A unique three-digit numeric
code assigned to each country or
territory according to ISO 3166.",
                                 "The region of the world where the
event took place.", "The country or territory in which
the event took place.",
                                 "The largest sub-national
administrative region in which the
event took place.",
                                 "The second largest sub-national administrative region in which the event took place.",
                                 "The third largest sub-national administrative region in which the event took place.",
                                 "The name of the location at which
the event took place.", "The latitude of the location in
four decimal degrees notation
(EPSG:3857).", "The longitude of the location in
four decimal degrees notation
(EPSG:3857).",
                                 "A numeric code between 1 and
3 indicating the level of certainty
of the location recorded for the
event. The higher the number,
the lower the precision.",
                                 "The sources used to record the
event. Separated by a semicolon.", "An indication of the geographic
closeness of the used sources to
the event", "A short description of the event.",
                                 "The number of reported fatalities
arising from an event. When there
are conflicting reports, the most
conservative estimate is recorded", "Additional structured information
about the event. Separated by a
semicolon.",
                                 "An automatically generated Unix timestamp that represents the exact date and time an event was last uploaded to the ACLED API."), check.names = F
)

returned_data_acled %>% 
  kable(align=rep('c')) %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(2, width = "20em") %>% 
  column_spec(3, width = "50em")

```

## ACLED Help

Rather than referring to this table during coding, and to provide a deeper understanding of a column’s meaning, you can use the  `acledR::acled_help()` function. This function will redirect you to the relevant documentation for the column in question. 

```{r,eval=FALSE}

acled_help(column = NULL)

```

The function is very straightforward, requiring only one argument.  `column` refers to the column for which you would like to be provided with more information. 

**Note**: `acled_help()` only supports one column at a time. 

For example, if you would like to know more about the *fatalities* column, you can run`acled_help(column="fatalities")`, prompting the fatalities vignette to appear in the RStudio Help pane.

While some columns have their own vignette, e.g. fatalities; others may be grouped together with documentation of similar columns, e.g. admins are in the geographical vignette.
