---
title: "Actors and Interactions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=F, message=F,
  echo = F,
  cache = T
)
library(acledR)
library(dplyr)
library(kableExtra)

```

See [the complete ACLED Codebook](https://acleddata.com/acleddatanew//wp-content/uploads/2021/11/ACLED_Codebook_v1_January-2021.pdf) for more details on each inter code and how they are named in the data. 


# Actors

ACLED recognizes a range of actors including state forces, rebels, militias, identity groups, demonstrators, civilians, and external and other forces. 

In ACLED, politically violent actors include government forces and its affiliates, rebel groups, militias, external or private forces (e.g. UN missions) and other political groups who interact over issues of political authority (e.g. territorial control, government control, access to resources, etc.). All organised actors have an official name and a political purpose, and use violence for political means. For inclusion as agents of political violence, organizations must be cohesive and not assembled for single events, with the exception of riots and protests. Further, the events of organizations must be connected to each other as a means to achieve a larger political purpose. This necessary and sufficient definition of actors allows for the establishment of campaigns and trajectories of movements. Protesters, rioters and civilians are actors whose inclusion deviates from the organization and armed group rules. Rioters and protesters involve spontaneous, atomic acts of organization that may, or may not, continue beyond a discrete event. Civilians are those who do not actively choose to be involved in an event. 

The name of each actor is noted in the `actor1` and `actor2` columns of the dataset.

The `assoc_actor_1` and `assoc_actor_2` columns record the associated groups for specific events as well as the identity of specific actors. In the former case, an associated group may be allies in actions, like two armed organized groups that are engaging in attacks against a common enemy. In the latter case, the `assoc_actor_1` and `assoc_actor_2` columns may record additional information concerning the victims of an attack or the socio-political affiliation of demonstrators or ethno-religious identity of a civilian victim. An ‘Explosions/Remote violence’ event that is intended for an armed, organized group, but also affects civilians, will have both groups noted – the primary actor will be the ostensibly intended target, while civilians (the collateral damage in this case) would be the associated actor.


# Interaction Codes

The type of group for each actor is referred to by "interaction codes". There are currently 8 categories of actors, each detailed in the table below. The type for `actor1` is recorded in the `inter1` column, while `inter2` codes the type for `actor2`. 

```{r}
tibble(`Interaction Code` = 1:8, 
       `Actor Category` =
         c("State Forces",
           "Rebel Groups",
           "Political Militias",
           "Identity Militias",
           "Rioters",
           "Protestors",
           "Civilians",
           "External/Other Forces"),
       Definition = c("Collective actors that are recognised to perform government functions, including military and police, over a given territory.", 
                      "Political organizations whose goal is to counter an established national governing regime by violent acts.",
                      "A more diverse set of violent actors, who are often created for a specific purpose or during a specific time period (i.e. Janjaweed largely active in Sudan) and for the furtherance of a political purpose by violence.",
                      "A broad category of “identity militias” that signifies armed and violent groups organized around a collective, common feature including community, ethnicity, region, religion or, in exceptional cases, livelihood.",
                      "Rioters are individuals or ‘mobs’ who either engage in violence during demonstrations or in spontaneous acts of disorganised violence.",
                      "Protesters are peaceful, unarmed demonstrators, though they may be targets of violence by other groups.",
                      "Civilians, in whatever number or association, are victims of violent acts within ACLED as they are, by definition, unarmed and, hence, vulnerable.",
                      "Small categories of “other” actors include international organisations, state forces active outside of their main country of operation, private security firms and their armed employees, and hired mercenaries acting independently.")) %>% 
  kable() %>% 
  kable_styling("striped", full_width = T)
```

# Joined Interaction Codes

The joined interaction code is the combination of the two `inter` codes associated with the two main actors. Single actor type codes are recorded in `inter1` and `inter2` columns, and the compounded number is recorded in the `interaction` column. For example, if a country’s military fights a political militia group, and the respective `inter1` and `inter2` codes are "1" and "3", respectively, the compounded Interaction is recorded as "13". 

Interaction numbers are always the smallest possible number (for example, 37 instead of 73), regardless of the order of `actor1` and `actor2`. Interaction codes are recorded for all events, including non-violent activity. For one-sided events, the empty second actor category is coded as "0". If a non-violent rebel event occurs where only `inter1` is noted with a "2", "20" is coded in the `interaction` column. Only the main actors recorded in the `actor1` and `actor2` columns are the basis for the interaction codes.

```{r}

inter1 <- c("Sole Military Action",
            "Military versus Military",
            "Military versus Rebels",
            "Military versus Political Militia",
            "Military versus Communal Militia",
            "Military versus Rioters",
            "Military versus Protestors",
            "Military versus Civilians",
            "Military versus Other")

inter2 <- c("Sole Rebel Action",
            "Rebels versus Rebels",
            "Rebels versus Political Militia",
            "Rebels versus Communal Militia",
            "Rebels versus Rioters",
            "Rebels versus Protestors",
            "Rebels versus Civilians",
            "Rebels versus Other")

inter3 <- c("Sole Political Militia Action",
            "Political Militia versus Political Militia",
            "Political Militia versus Communal Militia",
            "Political Militia versus Rioters",
            "Political Militia versus Protestors",
            "Political Militia versus Civilians",
            "Political Militia versus Other")

inter4 <- c("Sole Communal Militia Action",
            "Communal Militia versus Communal Militia",
            "Communal Militia versus Rioters",
            "Communal Militia versus Protestors",
            "Communal Militia versus Civilians",
            "Communal Militia versus Other")

inter5 <- c("Sole Rioter Action",
            "Rioters versus Rioters",
            "Rioters versus Protestors",
            "Rioters versus Civilians",
            "Rioters versus Other")

inter6 <- c("Sole Protestor Action",
            "Protestors versus Protestors",
            "Protestors versus Civilians",
            "Protestors versus Other",
            "Other Actor versus Civilians",
            "Sole other Action")

intertab <- tibble(interaction_code = c(10, 11, 12, 13, 14, 15, 16, 17, 18,
                            20, 22, 23, 24, 25, 26, 27, 28,
                            30, 33, 34, 35, 36, 37, 38, 
                            40, 44, 45, 46, 47, 48, 
                            50, 55, 56, 57, 58, 
                            60, 66, 67, 68, 78, 80),
       interaction_name = c(inter1, inter2, inter3, inter4, inter5, inter6))


t1 <- intertab[1:9,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped"), full_width = T) %>% 
  pack_rows("Military versus", 1, 9, label_row_css = "background-color: #0D647C; color: #fff;")
t2 <- intertab[10:17,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped"), full_width = T) %>% 
  pack_rows("Rebels versus", 1, 8, label_row_css = "background-color: #0D647C; color: #fff;")
t3 <- intertab[18:24,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped"), full_width = T) %>% 
  pack_rows("Communal Militia versus", 1, 6, label_row_css = "background-color: #0D647C; color: #fff;")
t4 <- intertab[25:30,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped")) %>% 
  pack_rows("Political Militia versus", 1, 5, label_row_css = "background-color: #0D647C; color: #fff;")
t5 <- intertab[31:35,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped"), full_width = T) %>% 
  pack_rows("Rioters versus", 1, 4, label_row_css = "background-color: #0D647C; color: #fff;")
t6 <- intertab[36:39,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped"), full_width = T) %>% 
  pack_rows("Protestors versus", 1, 3, label_row_css = "background-color: #0D647C; color: #fff;")
t7 <- intertab[40:41,] %>% 
  kable(col.names = NULL) %>% 
  kable_styling(c("striped"), full_width = T) %>% 
  pack_rows("Other versus", 1, 2, label_row_css = "background-color: #0D647C; color: #fff;")

knitr::kables(list(t1, t2)) %>% 
  kable_styling(font_size = 10)

knitr::kables(list(t3, t4)) %>% 
  kable_styling(font_size = 10)

knitr::kables(list(t5, t6)) %>% 
  kable_styling(font_size = 10)

knitr::kables(list(t7)) %>% 
  kable_styling(font_size = 10, full_width = F)





```

# `acledR` Example

As an example of the actors and interactions in ACLED, let's consider Yemen in January 2022 using the `acled_api()` function to call the API (read about conflict in Yemen at [ACLED's Yemen Research Hub](https://acleddata.com/middle-east/yemen/)):
  
```{r, echo = T}
library(acledR)
acled_access(email = Sys.getenv("email_address"), key = Sys.getenv("acled_key")) 
df_yemen <- acled_api(countries = c("Yemen"),
                      start_date = "2022-01-01",
                      end_date = "2022-01-31",
                      monadic = F, 
                   prompt = F)
```

During this month, there were 894 ACLED events, 28 unique `actor1` entries, and 24 unique `actor2` entries. 

```{r, echo = T}
# number of ACLED events
df_yemen$event_id_cnty %>% length()
# number of unique actor1s
df_yemen$actor1 %>% unique() %>% length()
# number of unique actor2s
df_yemen$actor2 %>% unique() %>% length()
```

The most common `actor1` was "Operation Restoring Hope", accounting for over 50% of events. 

```{r, echo = T}
df_yemen %>% 
  janitor::tabyl(actor1) %>% 
  janitor::adorn_rounding(2) %>% 
  arrange(-n) %>% 
  head(5)
```


We can see that "Operation Restoring Hope" is coded with `inter1 = 8`, indicating that this actor falls in the "External/Other Forces" category. 

```{r, echo = T}
df_yemen %>% 
  filter(actor1 == "Operation Restoring Hope") %>% 
  pull(inter1) %>% 
  unique()
```

The most common `actor2` was `NA` at 43%, indicating that no `actor2` was coded for these events. The next most frequent `actor2` was the Military Forces of Yemen. 

```{r, echo = T}
df_yemen %>% 
  janitor::tabyl(actor2) %>% 
  select(-valid_percent) %>% 
  arrange(-n) %>% 
  head(5)
```

Inspecting further, we see from the joined `interaction` column that most events were coded as "Sole other Action", mostly from the activity of "Operation Restoring Hope".

```{r, echo = T}
df_yemen %>% 
  janitor::tabyl(interaction) %>% 
  janitor::adorn_rounding(2) %>% 
  arrange(-n) %>% 
  head(5)

df_yemen %>% 
  filter(actor1 == "Operation Restoring Hope") %>% 
  janitor::tabyl(interaction) %>% 
  janitor::adorn_rounding(2) %>%
  arrange(-n) %>% 
  head(5)
```

Using the `sub_event_type` column in tandem with the `actor1` column, we see that "Operation Restoring Hope" mostly conducted air/drone strikes over this period. 

```{r}
df_yemen %>% 
  filter(actor1 == "Operation Restoring Hope") %>% 
  janitor::tabyl(sub_event_type) %>% 
  janitor::adorn_rounding(2) %>%
  arrange(-n) %>% 
  head(5)
```
