---
title: "Fatalities"
date: "2022-09-22"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fatalities

The study of fatalities in regions of conflict is one of the most
popular ways approaches of looking at conflict. However, while popular,
it is not always reliable. Tracking fatalities is often a complicated
subject within conflict studies. While it can provide important context
on escalations of violence, it can sometimes mislead audiences. Conflict
and political violence does not need to include fatalities for it to
devastate the life of people impacted by the conflict. Additionally,
fatalities are an elusive metric, as information about them can
sometimes be incomplete or erroneous, and if presented incorrectly, it
can lead to misinformation.

Yet, given the importance of tracking fatalities in a region of
conflict, ACLED puts a lot of effort on maintaining a transparent,
consistent and reliable methodology on how it tracks fatalities. While
this vignette briefly introduces such methodology, users are encouraged
to visit our [guide](https://acleddata.com/download/35143/) to get a
more comprehensive picture on how ACLED measures fatalities.

## How do we code fatalities?

ACLED's researchers are tasked with seeking more information about the
event to triangulate the numbers of a particular event reports, to avoid
over-relying in one source, mostly relaying on media and local partners
to gather information on fatalities of a given event. Furthermore,
researchers look for reports of **fatalities instead of casualties, as
casualties may also include injuries**. Thus, if researchers encounter
reports of casualties, ACLED counts it as 0 fatalities - unless there
are reports of deaths.

Given the fact that in almost any case reports over fatalities are
estimations, ACLED attempts to find the **most conservative available
estimate**, the lowest possible estimation of fatalities. While ACLED
acknowledges that this approach can lead to an underestimation of
fatalities, our efforts are to provide the best possible estimation,
even if lower, of fatalities rather than entirely arbitrary ranges also
built on assumptions.

As mentioned across ACLED's methodology, we relay on media and local
partners' reports to code fatalities. Because of that, if the reports do
not make any mention of fatalities, these are not going to be coded. In
some cases, however, sources may mention that fatalities occurred
without including an actual number. When such cases occur - and
researchers are unable to find any complementary information on the
number of such fatalities - ACLED relies in a series of rules to provide
an estimation of fatalities:

-   Is this a significant attack, in an active war zone? If yes,
    fatalities are estimated at 10.

    -   E.g. An airstrike is dropped on a building in Yemen, resulting
        in an unspecified number of fatalities. [Yemen is a war zone and
        an airstrike is likely to yield high casualties]

-   Is this an attack of more limited scope, in an active war zone? If
    yes, fatalities are estimated at 3.

    -   E.g. A bomb explodes at a checkpoint in Afghanistan, resulting
        in deaths. [Afghanistan is a war zone, yet checkpoints are often
        not manned by large numbers of people]

-   Is this an attack outside of a war zone? If yes, fatalities are
    estimated at 3.

    -   E.g. Inter communal violence in Guinea leads to a number of
        deaths. [Guinea is not a warzone, and inter communal violence
        tends to not result in mass fatalities]

-   Is this a significant attack outside of a war zone? If yes,
    fatalities are estimated at 10.

    -   E.g. A large-scale attack by Al Shabaab results in an
        unspecified number of fatalities in Kenya. [Kenya is not a war
        zone, yet this is a large-scale attack involving an actor in a
        neighboring war zone]

For that purpose, ACLED defines a country as a war zone if a single
group or coalition is attempting to violently challenge the legitimacy
and authority of the central government, and that same actor is active
in over 20% of the territory - an assessment done qualitatively.
Conflict zones are countries where there is organized political
violence, yet it has not reach the threshold of a war zone. Unrest
refers to protests and riots occurring in countries where there may not
be much conflict occurring, yet unrest is commonplace. Finally, disorder
is something which affects every country in ACLED coverage, which refers
to all political violence and demonstration events.

## Are these fatalities measures reliable?

As mentioned before, when we are coding fatalities we are speaking of
estimations. Thus, there is always some room for new information to
appear and change such measure. In such cases, ACLED modifies the events
in the data set to reflect such new information. At the same time,
ACLED's efforts to provide the most conservative estimate gives an
additional layer of reliability, as the measures are included only if
they are backed by sources in the region.

## How does ACLED deal with uncertainty?

The reason why refer to fatalities as **estimates**, is because there is
always some degree of uncertainty. Tracking and reporting fatalities in
an active conflict zone while the events are on-going is a multi-faceted
complicated issue, where some actors may opt not to report on losses or
on victims of their own attacks, or where civilians may not be readily
available to report on these fatalities. Hence, it is not as simple as
increasing the number of sources used to code these fatalities, because
in some cases it may not be enough, and in others there may not be
enough reliable sources given bias - bias either on **how** they report
or on **what** do they report.

In order to address these short comings, ACLED develops regionally
tailor-made approaches on finding more information on a given region.
Contrary to just increasing the number of sources (which may just add
even more biases and worsen the quality of information - quantity
doesn't ensure quality), ACLED aims to generate particular source
combinations to reflect the reality of disorder and fatalities counts in
each space. Generally, ACLED prioritizes local sources' reports, such as
local partners and sub national media, as these tend to avoid the
sensationalization of lethality that often occurs in English language
media.

## acledR Example:

In order to demonstrate the use of the fatalities columns, we will look
at fatalities in violence against civilians across 2021 and 2022 (the
first half until August).

```{r, echo=FALSE}
library(acledR)
library(tidyverse)
library(lubridate)
library(plotly)

acled_access(email = "acledexamples@gmail.com", key = "M3PWwg3DIdhHMuDiilp5") #  This is an example, you will need to input your credentials.
df_years <- acled_api(start_date = "2021-01-01",
                        end_date = "2022-08-31",
                        event_types = "Violence against civilians",
                        monadic = F,
                        acled_access = T,
                        prompt = FALSE)

```

Lets now explore fatalities within the data set.

```{r, echo = T}

df_years %>% 
  janitor::tabyl(fatalities) %>% 
  janitor::adorn_rounding(2) %>% 
  arrange(-n) %>% 
  head(5) 

summary(df_years$fatalities)

```

As we can see, our fatalities are for the most part concentrated in
between 0 and 1, with the average being of 1.21 fatalities per event.
That in itself it is a noticeable aspect, but we would like to go beyond
this. We will set these in categories of tenths.

```{r}

df_fatalities <- df_years %>%
  mutate(fatalities_bins = case_when(
    fatalities > 0 & fatalities <11 ~ "1 to 10",
    fatalities > 10 & fatalities <101 ~ "11 to 100",
    fatalities > 101 & fatalities <1001 ~ "101 to 1000",
    TRUE ~ "0"
  ))

df_fatalities %>%
  janitor::tabyl(fatalities_bins) %>%
  janitor::adorn_rounding(2)%>%
  arrange(-n)

```

Now that we have a better understanding on how our fatalities data set
is distributed, lets look at its evolution across 2021 and 2022.

```{r}

df_fatalities_monthly <- df_fatalities %>%
  group_by(Month = floor_date(ymd(event_date), "month")) %>%
  summarise(fatalities_sum = sum(fatalities)) %>%
  mutate(column = "fatalities") ## we create this column just to make a workaround in the tooltip in the plotly graph. 

g <- ggplot(df_fatalities_monthly, aes(x=Month,y=fatalities_sum)) +
  geom_line(color = "navyblue", aes(group = column ,text=paste("Sum of fatalities: ", fatalities_sum)))+
  labs(title = "Total of fatalities per month across 2021 and the first half of 2022",
       x = "Year-Month",
       y = "Sum of fatalities")

ggplotly(g, tooltip = c("x","text"))
```

From our graph we can observe a peak in the number of fatalities in
March 2022. Lets go slightly more in-depth to this date, lets begin by
exploring which are the countries that saw the highest difference
compared to the prior month.

```{r}
df_fatalities_monthly_country <- df_fatalities %>%
  group_by(Month = floor_date(ymd(event_date), "month"), country) %>%
  summarise(fatalities_sum = sum(fatalities)) %>%
  filter(Month < ymd("2022-04-01") & Month > ymd("2022-01-01")) %>%
  arrange(country) %>%
  group_by(country) %>%
  summarise(difference = diff(fatalities_sum))%>%
  arrange(-difference) %>%
  filter(difference != 0) ## We are filtering out those countries which didn't change to make a more polished graph. 

gp <- ggplot(df_fatalities_monthly_country, aes(y=reorder(country,difference),x=difference)) +
  geom_col()+
  labs(title = "Difference of total fatalities per country between \n February and March 2022\n",
       x= "Total difference of fatalities",
       y= "Countries")

ggplotly(gp)
```

Interestingly, when we explore the fatalities, we see Mali and Ukraine
as the top countries making contributions to the new peak in fatalities,
as these two were the countries with the biggest uptick in fatalities in
March 2022 compared to February. If a reader saw the peak from the
previous line chart, one may have swiftly explained it by the outbreak
of the Russian invasion into Ukraine in late February - however, this
would have missed the uptick in violence in Mali by Malian and Wagner
forces, see our
[analysis](%22https://acleddata.com/10-conflicts-to-worry-about-in-2022/sahel/mid-year-update/%22)
