---
title: "Automatic Report of `r params$country`"
output:
    html_document:
      toc: true
      fig_crop: no
      toc_float: true
    pdf_document:
      toc: true
      fig_crop: no
params:
  acled_email: email  
  acled_key: key   
  end_date: !r Sys.Date()
  country: country          
  printcode: TRUE 
  start_date: "1997-01-01"
  data: data
subtitle: "Country: `r params$country` - From: `r params$start_date` to `r params$end_date`"  

---

<style type="text/css">
.main-container {
  max-width: 1400px;
  margin-left: auto;
  margin-right: auto;
}
</style>


```{r loading what we need, echo=FALSE}
library(acledR)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(plotly)
library(viridis)
library(mapview)
library(stringr)
library(sf)

defaultW <- getOption("warn") 

options(warn = -1)


```

## 1. General statistics

```{r table of statistics, echo=FALSE}


table_data_source <- acled_transform(params$data, type = "source") %>%
  group_by(source) %>%
  summarise(frequency = n()) %>%
  arrange(-frequency)

unique_number_of_sources <- length(unique(table_data_source$source))

table_data_actor <- acled_transform(params$data, type = "full_actors")%>%
  group_by(actor) %>%
  summarise(frequency = n()) %>%
  arrange(-frequency)

unique_number_of_actors <- length(unique(table_data_actor$actor))


table_data <- params$data %>%
  mutate(event_date = as.character(event_date))%>%
  group_by(country) %>%
  summarise(total_events = n(),
            total_fatalities = sum(fatalities),
            eventful_days = n_distinct(event_date))
table_data$total_actors <- unique_number_of_actors
table_data$total_sources <- unique_number_of_sources

table_data <- table_data %>%
  rename(Country=country, "Total Events" = total_events, "Reported Fatalities" = total_fatalities, "Eventful Days" = eventful_days, "Total Actors" = total_actors, "Total Sources" = total_sources)

kbl(table_data) %>%
  kable_styling()

```

### 1.1 Sub-Event Types frequencies
```{r Sub-Event type,echo = FALSE, fig.align='right',fig.width=10,fig.height=7}
sub_event_table <- params$data %>%
  group_by(event_type,sub_event_type) %>%
  summarise(frequency = n()) %>%
  filter(event_type != "Strategic developments")

zu <-ggplot(sub_event_table,aes(x=frequency,fill=event_type,
                                text= c(
                                  paste0("Sub-Event type: ", sub_event_type,"\n",
                                         "Event Type: ",event_type, "\n",
                                         "Frequency: ", frequency))))+
      geom_bar(aes(y=reorder(sub_event_type,frequency)),stat="identity", size=10)+
      scale_x_continuous(trans="log10")+
      labs(title = "Magnitude of Sub-Event types in a log10 scale",
           x="Frequency in a 10 logarithmic scale",
           y="Sub-Event Types",
           fill = "Event Types")

ggplotly(zu, tooltip = "text")

```
### 1.2 Most Frequent Actors

```{r actor breakdown, echo = FALSE, fig.align='right',fig.width=10,fig.height=7}
fu <- ggplot(slice_max(table_data_actor,order_by = frequency,n=30),
             aes(frequency,reorder(actor,frequency),text=c(paste0("Actor: ", actor,"\n","Frequency: ", frequency))))+
  geom_bar(stat="identity", fill= "slateblue3")+
  theme(legend.position="none")+
  labs(title = "Top 30 most frequent actors in the dataset",
       x="Frequency",
       y="Actor")

ggplotly(fu, tooltip = "text")
```

### 1.3 Most Frequent Sources

```{r sources brealdown, echo = FALSE, fig.align='right',fig.width=10,fig.height=7}
su <- ggplot(slice_max(.data = table_data_source,order_by = frequency, n=30),aes(frequency,reorder(source,frequency),fill=source,text=c(paste0("Source: ", source,"\n","Frequency: ", frequency)))) +
  geom_bar(stat="identity",fill= "slateblue3")+
  theme(legend.position = "none")+
  labs(title = "Top 30 most frequent sources utilized in the dataset",
       x= "Frequency",
       y= "Source")
ggplotly(su,tooltip = "text")
```


## 2. Monthly average of events in `r params$country`

```{r moving avergeas, echo = FALSE, fig.align='right',fig.width=10,fig.height=7}

df1 <- params$data %>%
  generate_counts(.,
                  unit_id = "country",
                  time_id = "event_date",
                  time_target = "month")

# Generate 3 moving average of total events per month
df1m <-
  generate_movers(data = df1,
                  var = "total_events",
                  unit_id = "country",
                  time_id = "event_month",
                  slide_funs = "mean",
                  slide_periods = 3)

average_per_year <- df1 %>%
  mutate(year = lubridate::floor_date(event_month,unit = "year")) %>%
  group_by(year) %>%
  summarise(average_events_per_month = mean(total_events))

ha <- ggplot(df1m, aes(x=event_month, text = paste0("Month & Year: ",lubridate::format_ISO8601(event_month,precision = "ym"),"\n","3 months moving average: ", total_events_moving_mean_3, "\n", "Total number of events this month: ", total_events)))+
    geom_point(aes(y=total_events))+
    geom_line(aes(y=total_events_moving_mean_3, group=1, color = "3 Months Moving Average"))+
    # geom_line(data=average_per_year,aes(x=year,y=average_events_per_month), color = "green")+
    labs(title = paste0("Moving 3 months average of total events in ", params$country),
         x= "Month - Year",
         y= "Number of Events")+
     scale_colour_manual(name = 'Legend',
         values=c('3 Months Moving Average' = 'blue'))

ggplotly(ha, tooltip = "text")

```



## 3. Map of events distribution in the country per admin1 levels

```{r map of stuff,echo = FALSE, fig.align='right',fig.width=10,fig.height=7}


admin1_country <- raster::getData(name = "GADM", country = params$country, level = 1) %>%
  # convert to sf object
  st_as_sf()

country_counts <-
  params$data %>% 
  # full_join(select(admin1_country, NAME_1, NAME_0), by = c("admin1" = "NAME_1", "country" = "NAME_0")) %>% 
  group_by(admin1) %>% 
  tally() %>% 
  full_join(admin1_country, by = c("admin1" = "NAME_1")) %>% 
  mutate(n = case_when(
    is.na(n) ~ 0L,
    TRUE ~ n
  )) %>%
  sf::st_as_sf()

library(mapview)
mapviewOptions(fgb = FALSE) # needed when creating web pages
mapview(country_counts["n"], at = seq(1,2500,200))

```

## 4. Strategic Developments

Strategic developments are events coded by ACLED, not because of them being events of political violence, but because of these being events that may trigger future events of political violence. As such, it is important to know that these may not be sistematically coded the same way. However, Users may found it useful to know which are the latest strategic developments. 

### 4.1 Number of Strategic Developments
```{r strategic develop[ments counts}
strategic_devs_counts <- params$data %>%
  filter(event_type == "Strategic Development") %>%
  group_by(sub_event_type) %>%
  summarise("Number of Strategic Developments" = n())%>%
  ggplot(aes(x+sub_event_type,y="Number of Strategic Developments"))+
  geom_bar()+
  labs(title = "Strategic Development counts",
       x = "Type of Strategic Development",
       y = "Number of Events")
```

### 4.2 Stream of latest Strategic Developments

```{r Latest strategic developments}

strategic_devs <- params$data %>%
  filter(event_type == "Strategic Developments") %>%
  select(event_id_cnty, event_date,Sub_event_type,actor1,assoc_actor_1,actor2,assoc_actor_2, region, country, admin1,admin2,admin3,
         location,latitude,longitude,source,notes) %>%
  arrange(-event_date) %>%
  slice_head(n=10)
```



