---
title: "Using ACLED's API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ACLED's API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = F, 
  message = F,
  cache = F
  )
```
  
## Accessing the API
  
To use ACLED's [API](https://acleddata.com/api-documentation/getting-started), you must first register an account in [myACLED](https://acleddata.com/myacled). You can find more information about registering your account by visiting [myACLED's FAQs](https://acleddata.com/myacled-faqs). 
  
You will authenticate your API requests using your myACLED email and password. Unlike in older versions of `acledR` and the ACLED API, you will not need an API key to authenticate your requests. Instead, ACLED's API uses OAuth and `acledR` will handle OAuth tokens internally for you. We recommend storing the email and password associated with your ACLED account as environmental variables for easy and safe access in your scripts. 
  
To do so, run
```{r, eval = F}
file.edit(file.path("~", ".Renviron"))
```

which will open your `.Renviron` file. Once open, set:

```{r, eval = F}
email_address = "your_email"
acled_password = "your_password"
```

Then save the file and restart your R session. You can confirm that they have been properly stored by running `Sys.getenv("email_address")` to return the stored email address and `Sys.getenv("acled_password")` to return the stored ACLED key in the console. You can then enter your credentials within each API call like:

```{r, eval = F}
acled_api(
  email = Sys.getenv("email_address"),
  password = Sys.getenv("acled_password"),
  ...)
```


As an alternative, you can manually enter your credentials within each API call. While we do not recommend this approach because it is the least secure, it may be suitable for some use cases. You can use this option by following this example:

```{r, eval = F}
acled_api(
  email = "email_address", 
  password = "acled_password",
  ...)
```

You can also leave the password field empty to open an interactive prompt for typing in your password:
```{r, eval = F}
acled_api(
  email = "email_address", 
  ...)
```

Whether you set your email and password via environmental variables, manually, or via an interactive prompt, `acledR` will manage associated OAuth tokens for you via `acledR::acledR::acled_auth()`, which relies on `httr2::req_oauth_password()`.

## ACLED API

`acled_api()` is a function you can use to request and process ACLED API calls. The function takes the following arguments:

```{r, eval = F}
acled_api(email = NULL,
          password = NULL,
          country = NULL,
          regions = NULL,
          start_date = "1997-01-01",
          end_date = Sys.Date(),
          timestamp = NULL,
          event_types = NULL,
          population = "none",
          monadic = FALSE,
          inter_numeric = FALSE,
          ...)

```

## Parameters for the API

### Geographical filters

You can use the `country` and `regions` parameters to specify the locations from which you would like to request data. If both values are `NULL` or are not included, the API will return data for all countries and regions. If you would like to request data for multiple countries, you can do so by using a vector of country names (e.g., `c("Argentina","Spain","Bolivia")`). Similarly, you can request data from one or more regions by using either a vector of region names or numeric codes. `acledR::acled_countries` and `acledR::acled_regions` show the full lists of countries and regions available. Please visit ACLED’s [Knowledge Base](https://acleddata.com/article-categories/region-specific-methodology/) for region-specific methodology questions.

### Temporal filters

You can specify the date range for which you would like to receive data by using the `start_date` and `end_date` parameters, both of which require data in the "yyyy-mm-dd" format.

You can use the `timestamp` parameter to select data that were added or updated over a specific time period. Please keep in mind that `timestamp` indicates when the event was added or modified in ACLED’s dataset, meaning that an event that occurred far in the past (i.e., with an old *event date*) may still have a recent timestamp if it was recently updated.

In practice, the `timestamp` parameter is typically not used for analysis but is instead used to keep your own dataset up to date as changes are made to ACLED’s data. To learn more about how to keep your datasets up to date, visit the [Keeping your datasets up to date](https://acled.github.io/acledR/articles/acled_update.html) page for an acledR approach or this [guide](https://acleddata.com/download/35179/) more relevant to Excel or other spreadsheet tools.

### Additional filters

You can also use the `event_types` argument to filter to specific *event_types* in ACLED data. To do so, you should enter the *event_type* of interest as a string or as a vector of strings (e.g., `event_types = "Battles"` or `event_types = c("Battles", "Protests")`). For a description of all available *event_types* in ACLED’s dataset, please refer to [ACLED’s codebook](https://acleddata.com/download/2827/).

ACLED data defaults to a wide (or dyadic) format, where each row contains multiple actor columns, with those actors interacting during the event. However, you can request a long (or monadic) format using the `monadic` argument. By default, this argument is `FALSE`, meaning you will receive a dyadic version of the data. When `monadic=TRUE`, the function will return a monadic ("long-form") data frame with only one actor (based on *actor1* and *actor2*). For transforming your dataset from wide to long without using the API, or transforming it based on different sets of columns, visit `acled_transform_longer()`. For more information on the difference between our wide/dyadic and monadic/long datasets, please visit our [API guide](https://apidocs.acleddata.com/acled_endpoint.html#dyadic-versus-monadic-formats---export_type)

Finally, you can use the `population` argument to specify if you want to include the estimated population exposure columns. This argument takes three options, `none` which returns no extra columns, `best` which only returns the population_best column, or `full` which returns all the estimated population columns. For more information, visit our [Conflict Exposure piece](https://acleddata.com/conflict-exposure/).

The `...` parameter represents any other arguments you might want to include in your API query, such as *ISO* or *Interaction*. If you want to use these filters or others not included in the list of parameters described above, then you can write them as `&paramenter=value`. For instance, you might wish to include `&iso=4` at the end of the function. You can visit ACLED’s [API guide](https://apidocs.acleddata.com/acled_endpoint.html) to learn more about other valid parameters.



## Handling big API calls

As is common when executing API calls, handling large volumes of data requires some special consideration. In ACLED’s case, the base API uses pagination to address some of these issues, but pagination can be confusing for newer users (see our [API guide](https://apidocs.acleddata.com/generalities_section.html#file-size-limits-pagination) for a more detailed explanation). Fortunately, this package avoids this issue. Instead of manual pagination, the `acled_api()` function splits the call automatically.

`acled_api()` will first estimate how much data you are requesting. You will then be prompted with a message which includes the following:

-   The number of countries for which data is being requested,

-   The number of estimated events requested (based only on country and year, and NOT event type),

-   The number of API calls needed, based on an estimate of how big the call is,

-   A question asking whether given this information and the number of available API calls linked to your account – you would like to proceed with your API call.

## Example - Requesting data with `acled_api()`

Imagine you are interested in events from Argentina occurring between June 1-30 2022

```{r, eval = FALSE}
library(acledR)
library(dplyr)

#Note: This is simply an example–you will need to include your own credentials rather than the email and key placeholders that are included below.

df_ar <- acled_api(
  email = Sys.getenv("email_address"),
  password = Sys.getenv("acled_password"),
  country = c("Argentina"),
  start_date = "2022-06-01",
  end_date = "2022-06-30",
  monadic = FALSE)
```

```{r, echo=FALSE}
library(acledR)
library(dplyr)

df_ar <- acledR::acled_old_dummy

```

This returns a tibble that includes each ACLED event in Argentina during the specified period:

```{r}
head(df_ar, 5)
```

If you wanted data from both Brazil and Colombia, you would execute the following:

```{r, eval = F}
df_br_co <- acled_api(
  email = Sys.getenv("email_address"),
  password = Sys.getenv("acled_password"),
  country = c("Brazil", "Colombia"),
  start_date = "2022-01-01",
  end_date = "2022-12-01",
  monadic = FALSE)
```

If you are interested in events occurring over a larger area, it may be simpler to omit the `country` parameter and include a `regions` argument instead. You could also include an `event_type` argument to receive only a specific type of event:

```{r, eval = F}
df_sa <- acled_api(email = Sys.getenv("email_address"),
                   password = Sys.getenv("acled_password"),
                   regions = c("South America"),
                   start_date = "2022-01-01",
                   end_date = "2022-12-01",
                   event_type = "Protests",
                   monadic = FALSE)
```

You can use the *timestamp* column/filter to specify the dates from which you would like to receive new or updated data. You can include the argument as either a string ("yyyy-mm-dd") or a numeric Unix timestamp:

```{r, eval = F}
df_br_co <- acled_api(
  email = Sys.getenv("email_address"),
  password = Sys.getenv("acled_password"),
  country = c("Brazil", "Colombia"),
  start_date = "2022-01-01",
  end_date = "2022-12-01",
  monadic = FALSE,
  # timestamp = "2022-01-24" -> in the case of string
  timestamp = 1643056974, # -> in the case of a numeric Unix timestamp)
```


If you would like to include only one type of *interaction* (e.g., "Rioters versus Civilians (57)"), then you can add *interaction* code to the `...` argument:

```{r, eval = F}
df_sa <- acled_api(email = Sys.getenv("email_address"),
                   password = Sys.getenv("acled_password"),
                   country = c("Brazil", "Colombia"),
                   start_date = "2022-01-01",
                   end_date = "2022-12-01",
                   monadic = FALSE,
                   ... = "&interaction=57")
```

You could also request the monadic version of the data by setting `monadic = TRUE`:

```{r, eval = F}
df_sa_monadic <- acled_api(email = Sys.getenv("email_address"),
                           password = Sys.getenv("acled_password"),
                           regions = c("South America"),
                           start_date = "2022-01-01",
                           end_date = "2022-01-01",
                           monadic = TRUE)
```

