---
title: "Deletions API: Keeping your dataset up to date"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deletions API: Keeping your dataset up to date}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---
```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

defaultW <- getOption("warn") 

options(warn = -1) 

```

ACLED is a living dataset, which means changes happen regularly. 

Three types of changes are recurrent in ACLED's dataset:

-   New events
    -   These are the weekly addition of events, all are published with a new distinctive *event_id_cnty*. 
    
-   Updates of previously published events
    -   In some cases we modify published events, as new information may appear. For example, a particular actor has claimed responsibility over an attack, or the number of fatalities has changed. In these cases, the *event_id_cnty* remains the same, but the new information is updated and the old information overwritten - including the *timestamp*. 

-   Deletion of events
    -   In some cases, we have to delete previous events. This happens when new information surfaces indicating that the event no longer fits the scope of ACLED's dataset, or it hints that two events are the same and the events are then merged. In these cases, the event is removed, including the *event_id_cnty*.

For users with a personal ACLED file, it is advised to check regularly which events are no longer in the actual ACLED dataset, and which have been updated. Otherwise, users risk having an outdated version. In this section we will walk through how to keep your dataset updated with the `acledR` package. 

## Keeping check of updates - `acled_update()`

As we covered in the more detailed guide about [updating your dataset](https://acleddata.com/download/35179/), in some cases events are updated. When they are updated, their *event_id_cnty* remains the same but all the rest of the information gets updated and overwritten. 

Different from [deleted events](https://acled.github.io/ACLED-api-guide/deleted_endpoint.html), there is no separate API endpoint to check for updated events. When events are updated, their timestamp changes to the timestamp of the update. This means, that you can find the modified events by utilizing `acledR::acled_api()` and providing your latest timestamp (i.e. `max({your ACLED dataset}$timestamp))`) in the `timestamp` argument of the function. If you get duplicated events, these are likely to be updated events, which you should keep only the one with a higher timestamp. 

To simplify this process, `acledR` includes a function which makes the update for you by following the steps previously explained: `acled_update`. 

```{r, eval=F}
acled_update(
  df,
  countries = NULL,
  regions = NULL,
  event_types = NULL,
  acled_access = TRUE,
  email = NULL,
  key = NULL,
  deleted = TRUE,
  prompts = T)
```

The function is composed of the following arguments: 
-   `df`: The dataframe to update, it has to have the same structure as ACLED's dyadic dataframes (i.e. the result of `acled_api()`)
-   `start_date`: Along with `end_date` these determine the temporal range for your update. They default to the date range of `df`. 
-   `end_date`: Along with `start_date` these determine the temporal range for your update. They default to the date range of `df`. 
-   `countries`: In case of your dataset to include particular countries, you can add them here so that the function filter new events to those countries. 
-   `regions`: In case of your dataset to include particular regions, you can add them here so that the function filter new events to those regions. 
-   `event_types`: In case of your dataset to include particular event_types, you can add them here so that the function filter new events to those event_types.
-   `acled_acess`: If you have already utilized `acled_acess()`, you can set this option as TRUE (default) and you wont be required to input email or key. 
-   `email`: The email registered in [ACLED's Access Portal](https://developer.acleddata.com/). Not required if `acled_access = TRUE`. 
-   `key`: The key registered in [ACLED's Access Portal](https://developer.acleddata.com/). Not required if `acled_access = TRUE`.
-   `deleted`: If TRUE, this function will also deal with deleted events, utilizing [ACLED API's deleted endpoint](https://acled.github.io/acledR/articles/acled_deletions_api.html). 
-   `prompts`: Use this option to supress or allow prompts regarding your call. See `acled_api()`. 

`acled_update()` will utilize the dataset you provide to gather new events based on the timestamp, and add them to the original dataset, as well as remove any deleted events (if  `deleted = TRUE`). 

## Keeping track of deleted events - `acled_deletion_api()`

When events are deleted, however, their *event_id_cnty* also gets deleted. For that, users can employ `acled_deletion_api()` to gather the *event_id_cnty* of deleted events, which you can use to filter out these events from your dataset. While users can find an example below, we kindly suggest you to visit our [guide](https://acleddata.com/download/35179/) and a more specific guide about the `deletions` [endpoint](https://acled.github.io/ACLED-api-guide/deleted_endpoint.html) in our API.

```{r, eval=F}
acled_deletions_api(
  email=NULL,
  key=NULL,
  date_deleted=NULL,
  acled_access=TRUE # If acledR::acled_access() has been used before, leave as TRUE. 
)
```

The function is rather simple to use, requiring only 2 (4 if `acledR::acled_access()` is not used beforehand). It requires a date or `unix` timestamp from which to start gathering deleted events (`date_deleted`), and whether `acledR::acled_access()` has been utilized before. 

When executing the function, it will make an API call to the `deleted` endpoint of ACLED's API, and return a dataframe with two columns, one for the *event_id_cnty*, and another for the *timestamp*. Every event in this dataframe is a deleted event. 

## Examples

In this section we will show how to use the `acled_deletions_api()` and `acled_update` to keep your datasets updated. 

### Dataset n°1 - Keeping track of modified events

Load the first dataset:

```{r, eval = T, message=FALSE}
library(acledR)
library(lubridate)
library(dplyr)
```

```{r}
acled_access(email = "acledexamples@gmail.com", key = "M3PWwg3DIdhHMuDiilp5") #  This is an example, you will need to input your credentials.

argen_dummy_acled_file <- acledR::acled_old_dummy # Here is our old personal ACLED dataset
```

When was the last time you gathered your dataset? The snippet below checks that. 
```{r}
latest_timestamp_unix <- max(argen_dummy_acled_file$timestamp)

latest_timestamp <- as_datetime(latest_timestamp_unix) 
```

The dataset has not been updated since `r print(latest_timestamp)`, so you need to update it. To do so, you just have to utilize `acled_update()`. In order to keep the update within the dates of our dataset, you can ignore the `start_date` and `end_date` arguments. Similarly, we going to keep `deleted=TRUE` to also remove deleted events.  

```{r}

new_argen_dataset <- acled_update(argen_dummy_acled_file, 
                                  countries = "Argentina", 
                                  acled_access = T, 
                                  prompts = FALSE) 
```

Now your dataset captures modified and newly created events. 

### Dataset n°2 - Keeping track of deleted events

First, load the dataset.

```{r, eval = T}

dummy_acled_file <- acledR::acled_old_deletion_dummy # Here is our personal ACLED dataset
```

For future reference, you should remember the number of events in your dataset, so that you can have a quick glance on how many events were deleted. In this dataset there are  `r nrow(dummy_acled_file)` events. Now proceed to retrieve the deleted events using the highest timestamp.

```{r, eval = T}

deleted_events <- acled_deletions_api(date_deleted = max(dummy_acled_file$timestamp)) # We are using the largest Unix timestamp as that is the last time we gathered data. 

nrow(deleted_events)
```

As we can see `r nrow(deleted_events)` events have been removed since the last time this data was downloaded. That doesnt mean, however, that every event here is part of the old dataset. Just in case, you can use `dplyr::filter()` to exclude every event that is part of the deleted events dataset. 

```{r, eval = T}
updated_dataset <- dummy_acled_file %>%
  filter(!(event_id_cnty %in% deleted_events$event_id_cnty)) # We exclude every event_id_cnty that also appears in the deleted_events result

nrow(updated_dataset)
```

Now your dataset has only `r nrow(updated_dataset)` rows, thus `r nrow(dummy_acled_file) - nrow(updated_dataset)` events were deleted.

In these examples we have shown how you can keep your dataset updated with the help of `acledR`. While in both of our examples we tackled the modifications and deletions separately, normally, you should apply these simultaneously. 

Best of luck!

```{r, echo=F}
options(warn = defaultW)
```

